---
title: "03_nanoSIMS_processing"
editor: visual
format:
  html:
    toc: true
    code-fold: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

Clear the environment

```{r}
rm(list=ls())
```

# Load libraries

Load tidyverse and plotting libraries

```{r}
#| output: false
library(tidyverse) # CRAN v1.3.2
library(ggdist)    # CRAN v3.2.0
library(ggforce)   # CRAN v0.4.1
library(isotopia)  # [github::isoverse/isotopia] v0.5.8
library(scales)    # CRAN v1.2.1

source("source/calculate_mu.R")
source("source/calculate_sigma_mu.R")

```

# Import data

### Load nanoSIMS data

```{r}
#| output: false


data_path <- "data/nanoSIMS_tables/collected_tables/"


if (file.exists("cache/nanoSIMS_data.RDS")) {
  roi_data <- readRDS("cache/roi_data.RDS")
} else {
  tsv_file_list = list.files(path = data_path, pattern="*.tsv")
  
  roi_data <- tibble()
  roi_data <- roi_data[-1,] # Remove the garbage row of the tibble
  i = 0
  
  for (tsv in tsv_file_list) {
    temp <- read_tsv(paste0(data_path, tsv),
                     col_types = cols())
    temp <- temp %>% mutate(filename = tsv)
    roi_data <- bind_rows(roi_data, temp)
    i <- i + 1
    print(paste("File", i, "of", length(tsv_file_list), "loaded."))
  }
  saveRDS(roi_data, "cache/roi_data.RDS")
}

rm(temp) # remove the temp
```

### Load auxillary data

```{r}
# Amendment target concentration data
amendments <- readxl::read_excel("data/13C-amendment-concentrations.xlsx") |> 
  rename(spike.F13C = at13C,
         spike.uM = concentration.um)

# DIC concentration data
DIC <- readRDS(file = "data/Caro DIC Run1 March 6/om23_dic_run1.RDS")

# VOA data (via EMSL NMR)
voa_corrected_ctl <- readRDS("cache/VOA_corrected.RDS")
```

#### Define color palette and load theme

```{r}
source("source/color_palettes.R")
source("source/theme_om23.R")
```

# Add metadata

```{r}
roi_data <- roi_data |> 
  # remove the suffix, whichi s 23 characters long
  mutate(sample_id = str_sub(filename, end = -23)) |> 
  # add well info
  mutate(
    well = str_sub(sample_id, end = 4),
    # parse the depth as a 3 digit number
    depth = as.numeric(str_extract(sample_id, "\\d{2,3}")),
    amendment = case_when(
      str_detect(sample_id, "AC") ~ "acetate",
      str_detect(sample_id, "FORM") ~ "formate",
      str_detect(sample_id, "BIC") ~ "bicarbonate",
      str_detect(sample_id, "NEG") ~ "negative control",
    )
  ) |> 
  # add factional abundance
  mutate(at13C = Ratio_13Cx12C/ (1 + Ratio_13Cx12C))
```

# Plot 1: Investigate raw 13F trends

```{r}
neg_ctl_mean <- roi_data |> 
  filter(amendment == "negative control") |> 
  pull(at13C) |> 
  mean(na.rm = TRUE)

neg_ctl_sd <- roi_data |> 
  filter(amendment == "negative control") |> 
  pull(at13C) |> 
  sd(na.rm = TRUE)

custom_order = c("BA1B negative control", "BA1B formate", "BA1B bicarbonate", "BA1B acetate", "BA4A formate", "BA4A bicarbonate", "BA4A acetate")

# Plot it
p_ratio <- roi_data |> 
  mutate(
    condition = paste(well, amendment),
    condition = fct_recode(condition, !!!setNames(custom_order, custom_order))
  ) |> 
  filter(amendment != "negative control") |> 
  ggplot(
  aes(
    x = at13C,
    y = amendment,
    color = amendment,
    shape = as.factor(depth)
  )
) +
  geom_point(
    position = position_jitter(height = 0.25),
    alpha = 0.5
    ) +
  geom_vline(xintercept = neg_ctl_mean, linewidth = 0.5, color = "red", alpha = 0.5) +
  annotate(geom = "rect", 
    xmin = neg_ctl_mean - (2*neg_ctl_sd),
    xmax = neg_ctl_mean + (2*neg_ctl_sd),
    ymin = -Inf, ymax = Inf,
    color = NA,
    fill = "red",
    alpha = 0.25
  ) +
  scale_color_manual(values = amendment_palette) +
  facet_wrap(vars(depth, well)) +
  #ggforce::facet_zoom(xlim = c(0, 0.1)) +
  theme_bw() +
  labs(
    x = latex2exp::TeX("$^{13}F$"),
    y = "",
    shape = "Depth (mbs)"
  ) +
  theme(
    strip.background = element_rect(fill = "#e3e3e3", color = NA)
    )
p_ratio
```

# Rate calculations

Incubations were prepared in 50 mL serum vials that were acid washed and combusted prior to sampling, sealed with autoclaved butyl rubber stoppers. Carbon isotopes were added at 99.99 at. % enrichment to aim for concentrations of:

| Amendment       | BA3A  | BA4A   | BA1B   |
|-----------------|-------|--------|--------|
| 13C Formate     | 50µM  | 50µM   | 50µM   |
| 13C Acetate     | 50µM  | 50µM   | 50µM   |
| 13C Bicarbonate | 100µM | 1000µM | 6000µM |

Inspect the data:

```{r}
amendments |> knitr::kable()
```

We must account for isotopic dilution of the tracer from carbon pools native to the sample.

Before our work with EMSL, we used the following estimates of OA concentrations in the well, taking the mean values.

|  Well  | Formate (µM) | Acetate (µM) |
|:------:|:------------:|:------------:|
| NSHQ04 |     2.33     |     1.37     |
| NSHQ14 |     1.73     |     1.25     |
| WAB71  |     1.52     |    0.631     |

```{r}
formate_estimate.um <- mean(2.33, 1.73, 1.52)
acetate_estimate.um <- mean(1.37, 1.25, 0.631)

paste0("Formate: ", formate_estimate.um, "uM")
paste0("Acetate: ", acetate_estimate.um, "uM")
```

But we now have new data from PNNL EMSL's NMR facility:

```{r}
voa_corrected_ctl |> 
  filter(compound %in% c("Acetate", "Formate")) |> 
  filter(!is_control) |> 
  select(sample_id, compound, uM_corr) |> 
  pivot_wider(
    values_from = uM_corr, 
    names_from = compound
    ) |> 
  rename(Acetate_uM = Acetate,
         Formate_uM = Formate) |> 
  gt::gt() |> 
  gt::data_color(columns = Acetate_uM, palette = "Blues") |> 
  gt::data_color(columns = Formate_uM, palette = "Greens")
```

```{r}
voa_concs <- voa_corrected_ctl |> 
  filter(compound %in% c("Acetate", "Formate")) |> 
  filter(!is_control) |> 
  select(sample_id, compound, uM_corr) |> 
  rename(voa_conc_uM = uM_corr) |> 
  mutate(voa_F13C = 0.01105639) |> # 13F of VPDB (13R = 0.011180)
  mutate(amendment = stringr::str_to_lower(compound)) |> 
  select(-compound)
```

### Add DIC/OA data to nanoSIMS tibble

Make a slim version of the DIC data

```{r}
DIC_slim <- DIC |> 
  # select just the OM23 well data (remove standards, blanks)
  filter(type == "Falcon Tube", !is.na(well)) |>
  select(well, depth, DIC_uM, d13C.scale, d13C.error.S) |> 
  # convert d13C to at13C (fractional abundance)
  mutate(
    DIC.d13C = delta(`13C` = d13C.scale, major = "12C", ref = "VPDB"),
    DIC.d13C.error = delta(`13C` = d13C.error.S, major = "12C", ref = "VPDB"),
    DIC.R13C = to_ratio(DIC.d13C),
    DIC.R13C.error = to_ratio(DIC.d13C.error ),
    DIC.F13C = as.numeric(to_abundance(DIC.R13C)),
    DIC.F13C.error = as.numeric(to_abundance(DIC.R13C.error))
  ) |> 
  select(-c(d13C.scale, d13C.error.S))


```

### Mass balance the effective label strength

The calculation of isotopic abundances in pools derived from the combination of isotopically differing materials is defined as:

$$
m_{\Sigma}F_{\Sigma} = m_1F_1 + m_2F_2 + ...
$$

where the m terms represent molar quantities of the element of interest and the F terms represent fractional isotopic abundances.

In cases of isotope dilution, an isotopic spike is added to a sample and the mixture is analyzed. For the mixture, we can rearrange to:

$$
(m_x + m_k)F_\Sigma = m_xF_x + m_kF_k \\
F_\Sigma = \frac{m_xF_x + m_kF_k}{m_x + m_k}
$$

where k is the spike, and x is the natural abundance pool.

```{r}
calculate_effective <- function(mx, mk, fx, fk) {
  # mx and fx are natural abundance molarity and abundance
  # mk and fk are isotope spike molarity and abundance
  effective_concentration = (mx*fx + mk*fk) / (mx + mk)
  return(effective_concentration)
}

calculate_effective(mx = 50, mk = 50, fx = 1, fk = 0.5) # 0.75
calculate_effective(mx = 1, mk = 50, fx = 0.01, fk = 0.9999) # 0.9804
calculate_effective(mx = 2000, mk = 6000, fx = 0.01, fk = 0.9999) # 0.7524
```

So let's apply this to our samples. We then calculate single-cell growth rate as per Caro et al. 2023, Kopf et al. 2015.

## Calculate growth rates

Now we'll join our data tables of DIC data and amendment data.

1.  Join in the DIC data
2.  Join in the VOA data
3.  Join in the amendment data
4.  Calculate the effective concentration `F13C.label` that the cells saw, based on dilution of endogenous carbon sources
5.  calculate growth rate `mu.d`
6.  Define `active` cells as cells with 13C enrichment \> 3SD of the blank run. Set inactive cells mu.d to 0.

```{r}
# set activity threshold
F13C.activity_threshold <- neg_ctl_mean + (3 * neg_ctl_sd)


# join the tibbles
 growth_data <- roi_data |> 
  # rename sample_id column to coupon_id column
  rename(coupon_id = sample_id) |>
  # create new column that matches sample_id
  mutate(sample_id = paste0(well, "-", depth), .before = everything()) |>
  # add the DIC data
  left_join(DIC_slim, by = c("well", "depth")) |>
  # add the amendment data
  left_join(amendments, by = c("well", "amendment")) |> 
  # add the VOA data
  left_join(voa_concs, by = c("sample_id", "amendment")) |> 
 mutate(
    F13C.label = calculate_effective(
      # define the concentration of natural abundance C source:
      mx = case_when(
        amendment == "bicarbonate" ~ DIC_uM,
        amendment == "acetate" ~ voa_conc_uM,
        amendment == "formate" ~ voa_conc_uM
      ),
      # define the F13C of natural abundance C source:
      fx = case_when(
        amendment == "bicarbonate" ~ DIC.F13C,
        amendment == "acetate" ~ voa_F13C,
        amendment == "formate" ~ voa_F13C
      ),
      # molarity of the isotope spike:
      mk = spike.uM,
      fk = spike.F13C
    )
  ) |> 
   # calculate apparent growth rate
   mutate(
     mu.d = calculate_mu(
       a = 1, # assume assimilation efficiency of the carbon source is 1, ignoring fractionation effects because they are unknown
       F2 = at13C,
       F1 = neg_ctl_mean,
       FL = F13C.label,
       t2 = case_when(
         # add the incubation times from OM23 sample matrix .xlsx
         # START: 7 FEB 2023
         # END: 19 MAR 2023
         str_detect(sample_id, "BA1B") ~ 40,
         str_detect(sample_id, "BA4A") ~ 40,
         # END: 02 OCT 2023
         str_detect(sample_id, "BA3A-20") ~ 238,
         # END: 10 OCT 2023
         str_detect(sample_id, "BA3A-150-AC") ~ 246,
         str_detect(sample_id, "BA3A-150-BIC") ~ 246,
         str_detect(sample_id, "BA3A-270-AC") ~ 246,
         # END: 18 OCT 2023
         str_detect(sample_id, "BA3A-150-FORM") ~ 254,
         str_detect(sample_id, "BA3A-270-FORM") ~ 254,
         str_detect(sample_id, "BA3A-270-BIC") ~ 254,
       ),
       t1 = 0
     )
   ) |> 
   # if cell F13C did not exceed 2SD of the blank, set active = FALSE and set mu.d to zero
   mutate(
     active = if_else(
       at13C >= F13C.activity_threshold,
       true = TRUE,
       false = FALSE
     ),
     mu.d = if_else(
       active,
       true = mu.d,
       false = 0
     ),
    gen.d = log(2)/mu.d
   )
```

Let's inspect the effective label strengths of our incubations

```{r}
effective_label_summary <- growth_data |> 
  group_by(well, depth, amendment) |> 
  select(well, depth, amendment, F13C.label) |> 
  unique()

knitr::kable(effective_label_summary)
```

### Plot raw 13F trends with effective label

```{r}
p_13F <- growth_data |> 
  mutate(
    condition = paste(well, amendment),
    condition = fct_recode(condition, !!!setNames(custom_order, custom_order)),
    well = factor(well, levels = c("BA1B", "BA4A", "BA3A"))
  ) |> 
  filter(amendment != "negative control") |> 
  mutate(
    depth_str = factor(case_when(
      depth == 20 ~ "Shallow (20m)",
      depth == 150 ~ "Intermediate (150m)",
      depth == 250 ~ "Deep (250/270m)",
      depth == 270 ~ "Deep (250/270m)"
    ), levels = c("Shallow (20m)", "Intermediate (150m)","Deep (250/270m)")
  )) |> 
  ggplot(
  aes(
    x = at13C,
    y = amendment,
    color = amendment,
  )
) +
  # geom_point(
  #   position = position_jitter(height = 0.2),
  #   alpha = 0.5,
  #   size = 0.3
  # ) +
  geom_point(
    position = position_nudge(y = - 0.2),
    alpha = 0.25,
    shape = 124,
    size = 5
    ) +
  ggdist::stat_pointinterval(color = "black", point_interval = "mean_qi") +
  geom_vline(xintercept = neg_ctl_mean, linewidth = 0.5, color = "black") +
  geom_segment(aes(
    x = F13C.label, 
    xend = F13C.label, 
    y = as.numeric(factor(amendment)) - 0.35,
    yend = as.numeric(factor(amendment)) + 0.35,
    ),
    linewidth = 0.5) +
  annotate(geom = "rect", 
    xmin = neg_ctl_mean - (2*neg_ctl_sd),
    xmax = neg_ctl_mean + (2*neg_ctl_sd),
    ymin = -Inf, ymax = Inf,
    color = NA,
    fill = "black",
    alpha = 1
  ) +
  scale_color_manual(values = amendment_palette) +
  scale_x_continuous(labels = scales::label_percent()) +
  facet_grid(depth_str~well, scales = "free_x") +
  coord_cartesian(xlim = c(0, 1)) +
  labs(
    x = latex2exp::TeX("Cellular $^{13}F$ Enrichment (%)"),
    y = "",
    color = "Carbon Amendment"
  ) +
  theme_om23() +
  theme(
    legend.position = "None"
  )
  
p_13F

cowplot::save_plot(plot = p_13F, filename = "fig_output/p13F.pdf", base_height = 8, base_width = 12)
```

### Plot Growth Rates

```{r}
#| eval: false
p_growth_rate <- growth_data |> 
  mutate(
    condition = paste(well, amendment),
    condition = fct_recode(condition, !!!setNames(custom_order, custom_order))
  ) |> 
  filter(amendment != "negative control") |> 
  ggplot(
  aes(
    x = mu.d,
    y = amendment,
    color = amendment
  )
) +
    geom_point(
    position = position_nudge(y = - 0.2),
    alpha = 0.25,
    shape = 124,
    size = 5
    ) +
  ggdist::stat_pointinterval(color = "black", point_interval = "mean_qi") +
  scale_color_manual(values = amendment_palette) +
  coord_cartesian(xlim = c(0, 0.06)) +
  facet_wrap(vars(well, depth), scales = "free_x") +
  theme_bw() +
  labs(
    x = latex2exp::TeX("Growth rate $(day^{-1})$"),
    y = "",
    color = "Depth (mbs)",
    shape = "Depth (mbs)"
  ) +
  theme_om23() +
  theme(
    legend.position = "None"
  )
p_growth_rate

cowplot::save_plot(plot = p_growth_rate, filename = "fig_output/p_growth_rate.pdf",
                   base_height = 8, base_width = 12)
```

### Plot Generation Times

```{r}
#| eval: false
# summarize gentime for plotting
gen_data <- growth_data |> 
  group_by(well, depth, amendment) |> 
  summarize(
    mu.d = mean(mu.d, na.rm = TRUE)
  ) |> 
  mutate(mean_gen.d = log(2)/mu.d)

p_growth_rate_w_gen <- growth_data |> 
  mutate(
    condition = paste(well, amendment),
    condition = fct_recode(condition, !!!setNames(custom_order, custom_order))
  ) |> 
  filter(amendment != "negative control") |> 
  ggplot(
  aes(
    x = mu.d,
    y = amendment,
    color = amendment
  )
) +
    geom_point(
    position = position_nudge(y = - 0.2),
    alpha = 0.25,
    shape = 124,
    size = 5
    ) +
  geom_text(data = gen_data |> filter(amendment != "negative control"), 
            aes(label = paste0(round(mean_gen.d), " days"),
                x = mu.d + 0.01,
                y = as.numeric(factor(amendment)) + 0.2
                ),
            color = "black"
            ) +
  ggdist::stat_pointinterval(color = "black", point_interval = "mean_qi") +
  scale_color_manual(values = amendment_palette) +
  coord_cartesian(xlim = c(0, 0.06)) +
  facet_wrap(vars(well, depth), scales = "free_x") +
  theme_bw() +
  labs(
    x = latex2exp::TeX("Growth rate $(day^{-1})$"),
    y = "",
    color = "Depth (mbs)",
    shape = "Depth (mbs)"
  ) +
  theme_om23() +
  theme(
    legend.position = "None"
  )
p_growth_rate_w_gen

cowplot::save_plot(plot = p_growth_rate_w_gen, filename = "fig_output/p_growth_rate_w_gen.pdf",
                   base_height = 8, base_width = 12)
```

#### Plot 2.1 Updated growth rate

```{r}
# define a plotting helper function
plot_growth_rate <- function(well_choice, legend_choice = "None") {
  color_choice = case_when(
      well_choice == "BA1B" ~ 1,
      well_choice == "BA3A" ~ 3,
      well_choice == "BA4A" ~ 2,
    )
  
  
  # now plot!
  growth_data |> 
  filter(amendment != "negative control") |> # filter negative controls out
  # add "quantified" where sample is quantified if 3sigma > mean neg CTL
  mutate(quantified = if_else(
    at13C > (neg_ctl_mean + (3*neg_ctl_sd)),
    true = TRUE,
    false = FALSE)
    ) |>
  # uncomment this to remove samples that were not above background from plot
  # filter(quantified) |>
  filter(well == well_choice) |> 
  mutate(
    depth_str = paste0(depth, "m"),
    depth_str = forcats::fct_reorder(
      depth_str, depth, .desc = TRUE
      )
  ) |> 
  ggplot() +
  aes(
    x = mu.d,
    y = depth_str,
    color = amendment,
    shape = amendment
  ) +
  geom_point(
    data = function(df) df |> filter(!quantified),
    position = position_jitterdodge(
      jitter.width = 0.1,
      jitter.height = 0,
      dodge.width = 0.5
    ),
    alpha = 0.5,
    size = 1,
    color = "gray30"
  ) +  
  geom_point(
    data = function(df) df |> filter(quantified),
    position = position_jitterdodge(
      jitter.width = 0.1,
      jitter.height = 0,
      dodge.width = 0.5
    ),
    alpha = 0.5,
    size = 1,
  ) +
  ggdist::stat_pointinterval(
    aes(group = amendment), 
    position = position_dodge(0.5),
    point_size = 4,
    color = "black"
    ) +
  scale_color_manual(values = amendment_palette) + 
  #scale_x_continuous(breaks = c(0, 0.2, 0.4, 0.6, .8)) +
  coord_cartesian(xlim = c(0, 0.05)) +
  # scale_x_log10(
  #   labels = scales::trans_format("log10", 
  #   scales::math_format(10^.x)),
  #   breaks = c(1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1e0)
  #   ) +
  facet_wrap(vars(well)) +
  #annotation_logticks(sides = "bt") +
  theme_bw() +
  labs(
    x = latex2exp::TeX("Cell-specific growth rate $(day^{-1})$"),
    y = "",
    color = "Carbon Source",
    shape = "Carbon Source"
  ) +
  theme_om23() +
  theme(
    legend.position = legend_choice,
    axis.ticks.y = element_blank(),
    plot.margin = margin(0.25,0.25,0.25,0.25, "cm"),
    strip.background = element_rect(fill = well_palette[color_choice]),
    strip.text = element_text(margin = margin(0.2,0,0.2,0, "cm"))
  )
}

p_growth_legend <- cowplot::get_legend(plot_growth_rate("BA1B", legend = "right"))

p_growth_combined <- cowplot::plot_grid(
  plot_growth_rate("BA1B"),
  plot_growth_rate("BA4A"),
  plot_growth_rate("BA3A"),
  p_growth_legend,
  nrow = 1
)
p_growth_combined

# cowplot::save_plot(
#   plot = p_growth_combined,
#   filename = "fig_output/growth_combined.pdf",
#   base_width = 10, base_height = 5.5
#   )

```

#### Plot 2.2: Updated gen time

```{r}
growth_data_summary <- growth_data |> 
  filter(amendment != "negative control") |> # filter negative controls out
  # add "quantified" where sample is quantified if 3sigma > mean neg CTL
  mutate(
    quantified = if_else(
      at13C > (neg_ctl_mean + (3*neg_ctl_sd)),
      true = TRUE,
      false = FALSE)
    ) |>
  filter(quantified) |>
  mutate(
    depth_str = paste0(depth, "m"),
    depth_str = forcats::fct_reorder(
      depth_str, depth, .desc = TRUE
      )
  ) |>
    group_by(well, depth_str, amendment) |> 
    summarize(mean_mu.d = mean(mu.d, na.rm = TRUE),
              max_mu.d = max(mu.d, na.rm = TRUE),
              min_mu.d = min(mu.d, na.rm = TRUE)) |>
    mutate(mean_gen.d = log(2) / mean_mu.d,
           max_gen.d = log(2) / min_mu.d,
           min_gen.d = log(2) / max_mu.d) |> 
    ungroup()

# define a plotting helper function
plot_gen_time <- function(well_choice, legend_choice = "None") {
  color_choice = case_when(
      well_choice == "BA1B" ~ 1,
      well_choice == "BA3A" ~ 3,
      well_choice == "BA4A" ~ 2,
    )
  
  
  # now plot!
  growth_data |> 
  filter(amendment != "negative control") |> # filter negative controls out
  # add "quantified" where sample is quantified if 3sigma > mean neg CTL
  mutate(quantified = if_else(
    at13C > (neg_ctl_mean + (3*neg_ctl_sd)),
    true = TRUE,
    false = FALSE)
    ) |>
  # uncomment this to remove samples that were not above background from plot
  # filter(quantified) |>
  filter(well == well_choice) |> 
  mutate(
    depth_str = paste0(depth, "m"),
    depth_str = forcats::fct_reorder(
      depth_str, depth, .desc = TRUE
      )
  ) |> 
  ggplot() +
  aes(
    x = gen.d,
    y = depth_str,
    color = amendment,
    shape = amendment
  ) +
  geom_point(
    data = function(df) df |> filter(!quantified),
    position = position_jitterdodge(
      jitter.width = 0.1,
      jitter.height = 0,
      dodge.width = 0.5
    ),
    alpha = 0.5,
    size = 1,
    color = "gray30"
  ) +
  geom_point(
    data = function(df) df |> filter(quantified),
    position = position_jitterdodge(
      jitter.width = 0.1,
      jitter.height = 0,
      dodge.width = 0.5
    ),
    alpha = 0.5,
    size = 1,
  ) +
  geom_boxplot(
    color = "black",
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = 0.5),
    width = 0.25,
    size = 0.8
  ) +
  scale_color_manual(values = amendment_palette) + 
  coord_cartesian(xlim = c(1, 1e5)) +
  scale_x_log10(
    labels = scales::trans_format("log10",
    scales::math_format(10^.x)),
    breaks = c(1e0, 1e1, 1e2, 1e3, 1e4, 1e5)
    ) +
  facet_wrap(vars(well)) +
  annotation_logticks(sides = "bt") +
  theme_bw() +
  labs(
    x = latex2exp::TeX("Biomass Generation Time (days)"),
    y = "",
    color = "Carbon Source",
    shape = "Carbon Source"
  ) +
  theme_om23() +
  theme(
    legend.position = legend_choice,
    axis.ticks.y = element_blank(),
    plot.margin = margin(0.25,0.25,0.25,0.25, "cm"),
    strip.background = element_rect(fill = well_palette[color_choice]),
    strip.text = element_text(margin = margin(0.2,0,0.2,0, "cm"))
  )
}

p_gen_time <- cowplot::plot_grid(
  plot_gen_time("BA1B"),
  plot_gen_time("BA4A"),
  plot_gen_time("BA3A"),
  nrow = 1
)

p_gen_time

cowplot::save_plot(
  plot = p_gen_time,
  filename = "fig_output/gentime_combined.pdf",
  base_width = 10, base_height = 4
  )
```

#### Table: Gen Time Table

```{r}
growth_data_summary |> 
  select(well, depth_str, amendment, mean_gen.d, max_gen.d, min_gen.d) |> 
  mutate(depth_str = case_when(
      depth_str == "20m" ~ "Shallow (20m)",
      depth_str == "150m" ~ "Medium (150m)",
      depth_str == "250m" ~ "Deep (250 / 270m)",
      depth_str == "270m" ~ "Deep (250 / 270m)"
    )) |>
  mutate(depth_str = factor(depth_str, levels = c("Shallow (20m)", "Medium (150m)", "Deep (250 / 270m)"))) |> 
  gt::gt(rowname_col = "well", groupname_col = "depth_str")
    

```

## C assimilation rates

```{r}
# Define the mean and standard deviation of our natural abundance control
paste(
  "We set our F13C activity threshold as ",
  round(F13C.activity_threshold, 4),
  " or ",
  round(F13C.activity_threshold, 4) * 100,
  " at. % 13C"
  )

F13C.activity_threshold <- neg_ctl_mean + (2 * neg_ctl_sd)

paste0(
  "Our negative controls exhibit a F13C of ", round(neg_ctl_mean, 6), 
  "+/- ", round(2* neg_ctl_sd, 6), 
  " (2*SD). Therefore, our F13C activity threshold is ", 
  round(F13C.activity_threshold, 6), " or ",
  round(F13C.activity_threshold * 100, 4),
  " at. % 13C."
  )
```

The authors of [@stryhanyuk2018] define a term *K* as the fraction of the element incorporated by a cell during incubation with isotope-labeled growth substrates, relative to its initial cellular content. In other words, K represents the fraction of cellular Carbon that is from an isotope tracer, versus carbon *not* from an isotope tracer. The relative assimilation expressed with K values can be used to compare a relative difference and heterogeneity in single cell activity revealed in relative amount of assimilated material.

While [@stryhanyuk2018] define K using isotope ratio terms (*R*), we use fractional abundance notation where

$$
^{13}F = \frac{^{13}C}{^{13}C + ^{12}C}
$$

and

$$
^{13}F = \frac{^{13}R}{1 + \space ^{13}R}
$$

$$
K = \frac{aF_T - F_i}{aF_L - F_i}
$$

where:

-   $a$ is the assimilation efficiency fraction (0 - 1) of the target compound.

-   $F_T$ is the isotope fractional abundance at time of sampling (time = T)

-   $F_i$ is the initial isotope fractional abundance before the addition of tracer (or, a tracer-free negative control).

```{r}
calculate_K <- function(a = 1, FT, Fi, FL) {
  # a defaults to 1, assuming autotrophic growth
  # use other values as appropriate
  K = (a*FT - Fi) / (a*FL - Fi)
}
```

To calculate cell-specific assimilation rate, we can multiply the absolute value of an element mass per cell volume (for example, carbon partial density $\rho_C (g/\mu m^3)$ by the ROI-defined cellular volume ($V_i$ $\mu m^3$ ) calculated for each cell, to get the mass of carbon for each of the analyzed cells:

$$
m_i = \rho_C \times V_i
$$ where:

-   $m_i$ is the mass of carbon in a given cell

-   $\rho_C$ is the carbon partial density in a cell ($g/ \mu m^3$ )

-   $V_i$ is cellular volume ( $\mu m^3$)

Alternatively, as we do in this study, we can convert the cell-specific volume to cell-specific mass using the relationship described in [@khachikyan2019]:

$$
m_{carbon} = 197 \times V^{0.46}
$$

where:

-   $V$ refers to cellular volume in $\mu m^3$

-   $m_{carbon}$ refers to cell-specific carbon mass in femtograms (fg)

```{r}
# this function uses partial density of carbon
calculate_mcarbon_density <- function(rho, vol) {
  # rho is the carbon partial density of a cell in g/µm^3
  # v is cellular volume in µm^3
  mass = rho * vol
  return(mass)
}

# this function uses Khachikyan et al. 2019 empirically-derived relationship
calculate_mcarbon <- function(V) {
  mass = 197 * (V^0.46)
  return(mass)
}
```

In order for these calculations to work, we must have reasonable estimates of cellular volume that are derived from the ROI-area. [@trembath-reichert2021; @stryhanyuk2018] use the following relationship, which takes into account the volume of two capping hemispheres on a coccoidal or rod-shaped cell:

$$
V = \frac{1}{2}\pi \times W^2 \times \left( \frac{1}{3} W + \frac{1}{2}(L-W)\right)
$$

where:

-   L is cell length (major axis) in µm

-   W is cell width (minor axis) in µm

If L = W, this relationship will express the volume of a sphere, and so is applicable to coccoidal cells.

```{r}
calculate_volume <- function(L, W) {
  # L and W are major and minor axis lengths in um
  # volume is returned in um^3
  volume = ((1/2) * pi) * W^2 * ((1/3*W) + ((1/2)*(L-W)))
  return(volume)
}
```

To derive the mass of carbon assimilated by each cell, the cell-specific mass of carbon is multiplied by the fraction $K$ of carbon incorporated into the cell:

$$
\mu_i = m_i \times K
$$

where $µ_i$ is mass of assimilated carbon (*not to be confused with cell-specific growth rate, µ)*.

```{r}
calculate_assimilated_C <- function(mass, K) {
  # mass is cell-specific mass of carbon
  # K is fraction carbon that is incorporated
  mu = mass * K
  return(mu)
}
```

And the cell-specific assimilation rate of carbon ($r_c$) is calculated for each cell by dividing the assimilated carbon by time:

$$
r_C = \frac{u_i}{t} = \frac{\rho_C \times V_i \times K}{t}
$$

```{r}
calculate_assim_rate <- function(mu, t) {
  # mu is mass of assimilated carbon (fg)
  # the rate is assimilated carbon per unit time (fg/d for example)
  rate = mu / t
  return(rate)
}


calculate_assim_rate_density <- function(rho, vol, K, t) {
  # rho is the carbon partial density of a cell in g/µm^3
  # v is cellular volume in µm^3
  # K is the fraction of cellular carbon that is from assimilated tracer
  # t is change in time
  rate = (rho * vol * K) / t
}
```

The error in $r_C$ is calculated by accounting for uncertainties of the associated input values as follows:

$$
\Delta r_C = \sqrt{
\left( \frac{\partial r_c}{\partial \rho} \times \Delta \rho \right)^2 +
\left(\frac{\partial r_c}{\partial V_i} \times \Delta V  \right)^2 +
\left(\frac{\partial r_c}{\partial K} \times \Delta K  \right)^2 +
\left(\frac{\partial r_c}{\partial t} \times \Delta t  \right)^2
}
$$

and:

$$
\frac{\partial r_c}{\partial \rho} = \frac{V \times K}{t} \\
\frac{\partial r_c}{\partial V_i} = \frac{\rho \times K}{t} \\
\frac{\partial r_c}{\partial K} = \frac{\rho \times V}{t} \\
\frac{\partial r_c}{\partial t} = - \frac{\rho \times V \times K}{t^2}
$$

```{r}
calculate_assim_rate_error <- function(
    rho, # cell specific carbon density (g/µm^3)
    V, # cell specific volume (µm^3)
    K, # K
    t, # change in time
    del.rho, # uncertainty in rho
    del.V, # uncertainty in volume
    del.K, # uncertainty in K
    del.t # uncertainty in T
    ) {
  
  # define partial derivative terms
  dr.drho = (V*K) / t
  dr.dV = (rho * K) / t
  dr.dK = (rho * V) / t
  dr.dt = - (rho * V * K) / (t^2)
  
  del.rate = sqrt(
      (dr.drho * del.rho)^2 +
      (dr.dV * del.V)^2 +
      (dr.dK * del.K)^2 +
      (dr.dt * del.t)^2
  )
  
  return(del.rate)
}
```

```{r}
# how many microns per pixel?
# each image is 512 x 512 pixels
# each image is 25 µm x 25 µm area
um_per_px <- 25 / 512 # um/px

assim_data <- growth_data |> 
  # convert axis lengths to µm instead of pixels
  mutate(
    axis_major_length.um = axis_major_length * um_per_px,
    axis_minor_length.um = axis_minor_length * um_per_px
  ) |> 
  mutate(
    # add at13C natural abundance (F_i)
    at13C.t0 = neg_ctl_mean,
    # calculate K, fraction of cell assimilated from tracer
    K = calculate_K(a = 1, FT = at13C, Fi = at13C.t0, FL = F13C.label),
    # calculate cell volume
    vol = calculate_volume(L = axis_major_length.um, W = axis_minor_length.um),
    # calculate cell carbon mass in fg
    cell.C.fg = calculate_mcarbon(V = vol),
    # calculate mass of assimilated carbon
    assim.C.fg = calculate_assimilated_C(mass = cell.C.fg, K = K),
    # calculate rate of assimilation per day (34 day incubation)
    # fg/day
    rC.fg.d = calculate_assim_rate(mu = assim.C.fg, t = 34),
    # convert to fmol/cell/day by dividing fg / 12.01 fg/fmol
    rC.fmol.d = rC.fg.d / 12.01
  )


```

### Plot 3: Single Cell Carbon Assimilation Rates

```{r}
#| eval: false
assim_summary <- assim_data |> 
  group_by(well, depth, amendment) |> 
  summarize(
    rC.fmol.d = mean(rC.fmol.d)
  )


p_carbon_assim <- assim_data |> 
  filter(amendment != "negative control") |> # filter negative controls out
  # add "quantified" where sample is quantified if 3sigma above the mean natabund ctl
  mutate(quantified = if_else(
    at13C > (neg_ctl_mean + (3*neg_ctl_sd)),
    true = TRUE,
    false = FALSE
  )) |> 
  filter(quantified) |> 
  ggplot() +
  aes(
    x = rC.fmol.d,
    y = amendment,
    color = amendment
  ) +
  ggdist::stat_halfeye(color = "black", scale = 0.5) +
  geom_point(
    position = position_nudge(y = -0.2),
    alpha = 0.5,
    shape = 124,
    size = 3
    ) +
  geom_text(
    data = assim_summary |> filter(amendment != "negative control"),
    aes(
      x = rC.fmol.d + 0.1,
      y = as.numeric(factor(amendment)) + 0.2,
      label = round(rC.fmol.d, 4)
      ),
    color = "black"
  ) +
  scale_color_manual(values = amendment_palette) + 
  # scale_x_log10(
  #   breaks = trans_breaks("log10", function(x) 10^x),
  #   labels = trans_format("log10", math_format(10^.x))
  # ) +
  # annotation_logticks(sides = "b") +
  scale_x_continuous(
    labels = scales::trans_format("log10", 
    scales::math_format(10^.x))
    ) +
  coord_cartesian(xlim = c(0, .85)) +
  facet_wrap(vars(well, depth)) +
  theme_bw() +
  labs(
    x = latex2exp::TeX("fmol C $cell^{-1}$ $day^{-1}$"),
    y = "",
    color = "Depth (mbs)",
    shape = "Depth (mbs)"
  ) +
  theme_om23() +
  theme(
    legend.position = "None"
  )
    
p_carbon_assim

cowplot::save_plot(
  plot = p_carbon_assim,
  filename = "fig_output/carbon_assim_old.pdf",
  base_height = 7,
  base_width = 9
  )
```

#### Plot 3.1: Updated

```{r}
# define a plotting helper function
plot_carbon_assim <- function(well_choice) {
  color_choice = case_when(
      well_choice == "BA1B" ~ 1,
      well_choice == "BA3A" ~ 3,
      well_choice == "BA4A" ~ 2,
    )
  
  # now plot!
  assim_data |> 
  filter(amendment != "negative control") |> # filter negative controls out
  # add "quantified" where sample is quantified if 3sigma > mean neg CTL
  mutate(quantified = if_else(
    at13C > (neg_ctl_mean + (3*neg_ctl_sd)),
    true = TRUE,
    false = FALSE)
    ) |>
  # uncomment this to remove samples that were not above background from plot
  # filter(quantified) |>
  filter(well == well_choice) |> 
  mutate(
    depth_str = paste0(depth, "m"),
    depth_str = forcats::fct_reorder(
      depth_str, depth, .desc = TRUE
      )
  ) |> 
  ggplot() +
  aes(
    x = rC.fmol.d,
    y = depth_str,
    color = amendment,
    shape = amendment
  ) +
  geom_point(
    data = function(df) df |> filter(!quantified),
    position = position_jitterdodge(
      jitter.width = 0.1,
      jitter.height = 0,
      dodge.width = 0.5
    ),
    alpha = 0.5,
    size = 1,
    color = "gray30"
  ) +  
  geom_point(
    data = function(df) df |> filter(quantified),
    position = position_jitterdodge(
      jitter.width = 0.1,
      jitter.height = 0,
      dodge.width = 0.5
    ),
    alpha = 0.5,
    size = 1,
  ) +
  ggdist::stat_pointinterval(
    aes(group = amendment), 
    position = position_dodge(0.5),
    point_size = 4,
    color = "black"
    ) +
  scale_color_manual(values = amendment_palette) + 
  #scale_x_continuous(breaks = c(0, 0.2, 0.4, 0.6, .8)) +
  coord_cartesian(xlim = c(1e-5, 1e0)) +
  scale_x_log10(
    labels = scales::trans_format("log10", 
    scales::math_format(10^.x)),
    breaks = c(1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1e0)
    ) +
  facet_wrap(vars(well)) +
  annotation_logticks(sides = "bt") +
  theme_bw() +
  labs(
    x = latex2exp::TeX("fmol C $cell^{-1}$ $day^{-1}$"),
    y = "",
    color = "Carbon Source",
    shape = "Carbon Source"
  ) +
  theme_om23() +
  theme(
    legend.position = "None", # legend?
    axis.ticks.y = element_blank(),
    plot.margin = margin(0.25,0.25,0.25,0.25, "cm"),
    strip.background = element_rect(fill = well_palette[color_choice]),
    strip.text = element_text(margin = margin(0.2,0,0.2,0, "cm"))
  )
}

plot_carbon_assim("BA1B")

# get a legend
p_carbon_assim_ba4a_w_legend <- assim_data |> 
  filter(amendment != "negative control") |> # filter negative controls out
  # add "quantified" where sample is quantified if 3sigma above the mean natabund ctl
  mutate(quantified = if_else(
    at13C > (neg_ctl_mean + (3*neg_ctl_sd)),
    true = TRUE,
    false = FALSE
  )) |> 
  filter(quantified) |>
# choose the well:
  filter(well == "BA4A") |> 
  mutate(
    depth_str = paste0(depth, "m"),
    depth_str = forcats::fct_reorder(
      depth_str, depth, .desc = TRUE
      )
  ) |> 
  ggplot() +
  aes(
    x = rC.fmol.d,
    y = depth_str
  ) +
  geom_point(
    aes(
      color = amendment,
      shape = amendment
      ),
  position = position_jitterdodge(
    jitter.width = .1,
    dodge.width = .5
    ),
  alpha = 0.5,
  size = 2
  ) +
  ggdist::stat_halfeye(
    aes(group = amendment,
        ),
    position = position_dodge(0.5),
    point_size = 4,
    #color = "black"
    ) +
  scale_color_manual(values = amendment_palette) + 
  scale_x_continuous(breaks = c(0, 0.2, 0.4, 0.6, .8)) +
  coord_cartesian(xlim = c(0, .85)) +
  facet_wrap(vars(well)) +
  theme_bw() +
  labs(
    x = latex2exp::TeX("fmol C $cell^{-1}$ $day^{-1}$"),
    y = "",
    color = "Carbon Source",
    shape = "Carbon Source"
  ) +
  theme_om23() +
  theme(
    strip.background = element_rect(fill = well_palette[3]),
    strip.text = element_text(margin = margin(0.2,0,0.2,0, "cm"))
  )
p_carbon_assim_ba4a_w_legend

p_carbon_assim_legend <- cowplot::get_legend(p_carbon_assim_ba4a_w_legend)

p_carbon_assim_combined <- cowplot::plot_grid(
  cowplot::plot_grid(
    plot_carbon_assim(well_choice = "BA1B"),
    plot_carbon_assim(well_choice = "BA4A"),
    plot_carbon_assim(well_choice = "BA3A"),
    ncol = 3,
    nrow = 1
  ),
  p_carbon_assim_legend,
  rel_widths = c(0.9, 0.1)
) 
  

p_carbon_assim_combined

cowplot::save_plot(
  filename = "fig_output/carbon_assim_combined.pdf",
  plot = p_carbon_assim_combined, 
  base_width = 10, base_height = 4,
  )
```

### Table summary of C assimilation rates

Now we output the results of the above plot in table form

```{r}
assim_data_fmol_summary <- assim_data |> 
  group_by(well, depth, amendment) |> 
  filter(amendment != "negative_control") |> 
  summarize(
    mean_rC.fmol.d = mean(rC.fmol.d)
  ) |> 
  pivot_wider(
              names_from = "amendment", 
              values_from = mean_rC.fmol.d) |> 
  select(-`negative control`) |> 
  mutate(site = paste0(well, "-", depth, "m")) |> 
  ungroup()

assim_data_fmol_summary |> 
  select(-c(well, depth)) |> 
  gt::gt(rowname_col = "site") |> 
  gt::tab_header(title = "Carbon Assimilation Rates") |> 
  gt::tab_footnote(footnote = "All values reported in cell-specific fmol C per day.",
                   locations = gt::cells_column_labels(
                     columns = c(acetate, bicarbonate, formate))) |> 
  gt::tab_row_group(label = "BA1B", rows = c("BA1B-20m", "BA1B-150m", "BA1B-250m")) |> 
  gt::tab_row_group(label = "BA4A", rows = c("BA4A-20m", "BA4A-150m", "BA4A-270m")) |> 
  gt::tab_row_group(label = "BA3A", rows = c("BA3A-20m", "BA3A-150m", "BA3A-270m")) |> 
  gt::tab_spanner(columns = everything(), label = "Amendment") |> 
  gt::fmt_scientific(columns = c(acetate, bicarbonate, formate)) |> 
  gt::data_color(columns = c(acetate, bicarbonate, formate), palette = "Blues") 
```

### Table Summary of % Active Cells

```{r}
fraction_active_summary <- assim_data |> mutate(
  quantified = if_else(
    at13C > (neg_ctl_mean + (3*neg_ctl_sd)),
    true = TRUE,
    false = FALSE
    )
  ) |>
  group_by(
    well, depth, amendment
  ) |> 
  summarize(
    sumQ = sum(quantified),
    total = n()
  ) |> 
  mutate(fraction_active = sumQ / total)

  
  fraction_active_summary |> 
    ungroup() |> 
    filter(amendment != "negative control") |> 
    ggplot() +
    aes(
      x = amendment,
      y = fraction_active,
    ) +
    geom_point() +
    geom_segment(aes(xend = amendment), yend = 0) +
    scale_y_continuous(labels = label_percent()) +
    coord_polar() +
    facet_wrap(vars(depth, well)) +
    theme_bw()
  
  fraction_active_summary |> 
    select(well, depth, fraction_active, amendment) |> 
    mutate(depth = case_when(
      depth == 20 ~ "Shallow (20m)",
      depth == 150 ~ "Medium (150m)",
      depth == 250 ~ "Deep (250 / 270m)",
      depth == 270 ~ "Deep (250 / 270m)"
    )) |>
    filter(amendment != "negative control") |> 
    pivot_wider(values_from = fraction_active, names_from = well) |> 
    select(amendment, depth, BA1B, BA4A, BA3A) |> 
    gt::gt(rowname_col = "amendment", groupname_col = "depth") |> 
    gt::tab_header(title = "Fraction of Active Cells") |>
    gt::tab_footnote(
    footnote = "Where 'active' is defined as isotopic enrichment exceeding 3 standard deviations from the mean of the negative control.",
    locations = gt::cells_title()
  ) |> 
    gt::data_color(columns = BA1B, palette = "Reds") |> 
    gt::data_color(columns = BA4A, palette = "Blues") |> 
    gt::data_color(columns = BA3A, palette = "Greens") |> 
    gt::fmt_percent(columns = c(BA1B, BA4A, BA3A)) |> 
    gt::cols_align(align = "center")

  
```

### Plot 4: Single cell K

```{r}
assim_data |> 
  filter(amendment != "negative control") |> # filter negative controls out
  # add "quantified" where sample is quantified if 3sigma above the mean natabund ctl
  mutate(quantified = if_else(
    at13C > (neg_ctl_mean + (3*neg_ctl_sd)),
    true = TRUE,
    false = FALSE
  )) |> 
  filter(quantified) |> 
  ggplot() +
  aes(
    x = K,
    y = amendment,
    color = amendment
  ) +
  ggdist::stat_halfeye(color = "black", scale = 0.5) +
  geom_point(
    position = position_nudge(y = -0.2),
    alpha = 0.5,
    shape = 124,
    size = 3
    ) +
  scale_color_manual(values = amendment_palette) + 
  facet_wrap(vars(well, depth)) +
  theme_bw() +
  labs(
    x = "K",
    y = "",
    color = "Depth (mbs)",
    shape = "Depth (mbs)"
  ) +
  theme_om23() +
  theme(
    legend.position = "None"
  )
```

### Plot 5: Single Cell Volumes

```{r}
cell_vol_summary <- assim_data |> 
  group_by(well, depth, amendment) |> 
  summarize(
    vol = mean(vol, na.rm = TRUE)
  )

assim_data |> 
  filter(amendment != "negative control") |> # filter negative controls out
  # add "quantified" where sample is quantified if 3sigma above the mean natabund ctl
  mutate(quantified = if_else(
    at13C > (neg_ctl_mean + (3*neg_ctl_sd)),
    true = TRUE,
    false = FALSE
  )) |> 
  filter(quantified) |> 
  ggplot() +
  aes(
    x = vol,
    y = amendment,
    color = amendment
  ) +
  ggdist::stat_halfeye(color = "black", scale = 0.5, point_interval = "mean_qi") +
  geom_point(
    position = position_nudge(y = -0.2),
    alpha = 0.5,
    shape = 124,
    size = 3
    ) +
  geom_text(
    data = cell_vol_summary |> filter(amendment != "negative control"),
    aes(
      x = vol + 0.1,
      y = as.numeric(factor(amendment)) + 0.2,
      label = round(vol, 2)
      ),
    color = "black"
  ) +
  scale_color_manual(values = amendment_palette) + 
  scale_x_continuous(limits = c(0, 1)) +
  facet_wrap(vars(well, depth)) +
  theme_bw() +
  labs(
    x = latex2exp::TeX("Cell volume ($\\mu m ^3$)"),
    y = "",
    color = "Depth (mbs)",
    shape = "Depth (mbs)"
  ) +
  theme_om23() +
  theme(
    legend.position = "None"
  )
```

# How many cells?

How many cells did we analyze? Each cell analyzed is represented by a single ROI. Let's count the number of ROIs.

```{r}
n_cells <- roi_data |> 
  summarize(n()) |> 
  pull()

paste("A total of", n_cells, "individual microbial cells were analyzed by nanoSIMS in this study.")
```

# Export to cache

```{r}
# growth rate data
saveRDS(growth_data, "cache/SIMS_growth_data.RDS")
saveRDS(assim_data, "cache/SIMS_assim_data.RDS")
saveRDS(growth_data_summary, "cache/SIMS_growth_data_summary.RDS")
```
