---
title: "14b: Monte Carlo Input Distributions"
format: html
editor: source
editor_options:
  chunk_output_type: console
---

In R2R, requested that we specify input distributions. This script reads `cache/mc_input_params.RDS` the exact parameter objects (`PARAMS`, `assim_params`, `cell_density_params`) used by the Monte Carlo sampler in `14_power_calcs.qmd`

If `14_power_calcs.qmd` changes what it samples, re-run it first (it re-writes `cache/mc_input_params.RDS` as its final step before the sampler runs) and then re-run this script!

# Setup

```{r}
library(tidyverse)

mc_inputs <- read_rds("cache/mc_input_params.RDS")
n_sim <- mc_inputs$n_sim
global_params <- mc_inputs$global_params
assim_params <- mc_inputs$assim_params
cell_density_params <- mc_inputs$cell_density_params
```

The Monte Carlo simulation draws `r n_sim` independent samples per well/depth site (9 sites total = `r n_sim * 9` total simulations) from six input distributions. Three are global and three are site-specific (a distinct log-normal distribution fit per well or per well/depth).

# Global input distributions

`delta_G`, `hab_depth_km`, and `stim_factor` are each drawn from a Uniform distribution, identical for every well/depth site.

```{r}
global_dist_table <- tibble(
  parameter = c("delta_G", "hab_depth_km", "stim_factor"),
  description = c(
    "Gibbs free energy yield of hydrogenotrophic methanogenesis (4H2 + CO2 -> CH4 + 2H2O), bounded by two thermodynamic endmembers spanning the range of Samail ophiolite fluid geochemistry",
    "Habitable subsurface depth over which volume-specific power is integrated to landscape-scale flux",
    "H2-stimulation correction factor: divides MSP to account for ex situ mesocosm incubations being initiated at H2 concentrations above ambient, per the Monod-kinetics argument in 'Accounting for uncertainties'"
  ),
  distribution = "Uniform",
  min = map_dbl(
    global_params[c("delta_G", "hab_depth_km", "stim_factor")],
    "min"
  ),
  max = map_dbl(
    global_params[c("delta_G", "hab_depth_km", "stim_factor")],
    "max"
  ),
  units = c(
    "J / tonne H2 (1.26e10-2.393e10 J/tonne H2 = -25.1 to -47.8 kJ/mol H2)",
    "km",
    "dimensionless (divisor applied to MSP)"
  ),
  n_draws_per_site = n_sim
)

global_dist_table
```

Three additional constants set the *fallback* standard deviation (in log10 space) used for a site-specific log-normal distribution below, only when that site/well has fewer than 2 replicate measurements to estimate an empirical SD from:

```{r}
fallback_sd_table <- tibble(
  parameter = c(
    "cell_density_log_sd_default",
    "msp_log_sd_default",
    "cell_c_log_sd_default"
  ),
  applies_to = c(
    "cell_dens_mean_log10 (per well)",
    "MSP_mean_log10 (per well/depth)",
    "cellC_mean_log10 (per well/depth)"
  ),
  fallback_sd_log10 = map_dbl(
    global_params[c(
      "cell_density_log_sd_default",
      "msp_log_sd_default",
      "cell_c_log_sd_default"
    )],
    identity
  )
)

fallback_sd_table
```

# Site-specific input distributions

## Cell density (per well)

Log-normally distributed in cells/cm3 (i.e. `10^Normal(mean_log10, sd_log10)`), fit from replicate cell-density measurements per well (Templeton et al. 2021 data extracted by `27_convert_pdf_to_xl.qmd`). 

Rock density is assumed to be `r 2.7` g/cm3 to convert cells/g to cells/cm3.

```{r}
cell_density_dist_table <- cell_density_params |>
  mutate(
    distribution = "Log-normal (log10 cells/cm3)",
    median_cells_per_cm3 = 10^cell_dens_mean_log10,
    sd_is_fallback_default = n_replicates < 2
  ) |>
  select(
    well,
    distribution,
    mean_log10 = cell_dens_mean_log10,
    sd_log10 = cell_dens_sd_log10,
    median_cells_per_cm3,
    n_replicates,
    sd_is_fallback_default
  )

cell_density_dist_table
```

## Mass-specific power & single-cell carbon (per well/depth)

Both log-normally distributed, fit per well/depth site from the single-cell nanoSIMS-SIP dataset `cache/SIMS_assim_data.RDS`.

```{r}
site_dist_table <- assim_params |>
  mutate(
    MSP_distribution = "Log-normal (log10 W (g C)-1)",
    MSP_median_W_gC = 10^MSP_mean_log10,
    MSP_sd_is_fallback_default = n_cells < 2,
    cellC_distribution = "Log-normal (log10 fg C cell-1)",
    cellC_median_fgC = 10^cellC_mean_log10,
    cellC_sd_is_fallback_default = n_cells < 2
  ) |>
  select(
    well,
    depth,
    n_cells,
    MSP_distribution,
    MSP_mean_log10,
    MSP_sd_log10,
    MSP_median_W_gC,
    MSP_sd_is_fallback_default,
    cellC_distribution,
    cellC_mean_log10,
    cellC_sd_log10,
    cellC_median_fgC,
    cellC_sd_is_fallback_default
  )

site_dist_table
```

No site in this dataset has fewer than 2 quantified cells or fewer than 2 cell-density replicates, so the fallback default SDs above are not actually invoked for the results in the manuscript!

```{r}
stopifnot(!any(cell_density_dist_table$sd_is_fallback_default))
stopifnot(!any(site_dist_table$MSP_sd_is_fallback_default))
stopifnot(!any(site_dist_table$cellC_sd_is_fallback_default))
```

# Export

Write every table above to a single multi-sheet workbook.

```{r}
readme <- tibble(
  section = c(
    "Overview",
    "Global distributions",
    "Cell density",
    "MSP / cell carbon",
    "Sampling",
    "Source"
  ),
  notes = c(
    "The Monte Carlo simulation in 14_power_calcs.qmd (section '## run montecarlo') draws parameters from 6 input distributions per well/depth site: 3 global (identical at every site) and 3 site-specific (fit from data). See the 'global_distributions', 'cell_density_by_well', and 'MSP_and_cellC_by_site' sheets.",
    "delta_G, hab_depth_km, and stim_factor are Uniform, identical for all 9 sites.",
    "Cell density is Log-normal, fit per well (3 wells) from replicate cell-density counts (Templeton et al. 2021).",
    "Mass-specific power (MSP) and single-cell carbon content are each Log-normal, fit per well/depth (9 sites) from the nanoSIMS-SIP single-cell dataset.",
    paste0(
      n_sim,
      " independent draws per site (",
      n_sim * 9,
      " total simulations)."
    ),
    "Generated from cache/mc_input_params.RDS, written by 14_power_calcs.qmd."
  )
)

writexl::write_xlsx(
  list(
    readme = readme,
    global_distributions = global_dist_table,
    fallback_sd_defaults = fallback_sd_table,
    cell_density_by_well = cell_density_dist_table,
    MSP_and_cellC_by_site = site_dist_table
  ),
  path = "data_output/monte_carlo_input_distributions.xlsx"
)
```



```{r}
global_dist_table |> glimpse()
```